Computing Direct Sum Decompositions

### 2025-04-28 paper/main.tex
with any luck, everything is correct

We will refer to the collection of $k$-eigenvalues of $B_i$ as the ``eigenvalues of $\phi$''. Further, let $\mu_j \coloneqq \mu(\lambda_j)$ be the sum of the multiplicities of $\lambda_j$ in each $B_i$. Note that the multiplicity of $\lambda_j$ in $\chi_\phi(\lambda)$ is the sum of the multiplicities of $\lambda_j$ in each $\chi_{B_i}(\lambda)$.

@@

\pagebreak
\begin{prop}\label{prop:split-surj}
  Let $\lambda_1,\dots,\lambda_r$ be the eigenvalues of $\phi$ over $k$ and set $\psi_j = (\phi - \lambda_j\id_M)^{\mu_j}$ and $\psi = \psi_1\circ\cdots\circ\psi_r$. Then $M$ has a direct sum decomposition
  \( M = \ker\psi_1 \oplus \cdots \oplus \ker\psi_r \oplus \im\psi. \)
\end{prop}
\begin{proof}
  First, suppose $\lambda_{r+1},\dots,\lambda_t$ are the other eigenvalues of $\phi$ over the algebraic closure $\bar k$ and set $\rho = \psi_{r+1}\circ\dots\circ\psi_t$, then by the Cayley--Hamilton theorem $\chi_\phi(\phi) = \psi_1\circ\dots\circ\psi_r\circ\rho = \psi\circ\rho = 0$. Since the factors in $\chi_\phi(\phi)$ can be reordered, we can write $0 = \psi\circ\rho = \rho \circ \psi$. In particular, suppose the desired direct sum decomposition existed over $\bar k$, then we have \mahrud{WHY??}
  \begin{align*}
    M = \ker\psi_1\oplus\dots\oplus\ker\psi_r\oplus(\ker\psi_{r+1}\oplus\dots\oplus\ker\psi_t)
    &= \ker\psi_1\oplus\dots\oplus\ker\psi_r\oplus\ker\rho \\
    &= \ker\psi_1\oplus\dots\oplus\ker\psi_r\oplus\im\psi.
  \end{align*}
  Thus it suffices to prove the direct sum decomposition only over the algebraic closure.

  If $r = 1$ then $\im\psi_1=0$ is zero, hence we get the trivial decomposition $M = M \oplus 0$.

  %  We will find a right inverse to $\psi_1$ and $\psi_2$ using $\chi_\phi(\phi) = 0$
  If $r = 2$, let $A\colon M/\m M\to M/\m M$ be the $k$-linear map induced by $\phi$ and $A_i\coloneqq(A - \lambda_i)^{\mu_i}$ the map induced by $\psi_i$ for $i=1,2$. Observe that
  \begin{align*}
  \chi_\phi(\phi) = (\phi - \lambda_1\id_M)^{\mu_1} (\phi - \lambda_2\id_M)^{\mu_2}
  &= \left(\phi^{\mu_1} - \dots + c_1\phi - c_0\id_M\right)\circ\psi_2 \\
  &= \left(\phi^{\mu_1} - \dots + c_1\phi\right)\circ\psi_2 - c_0\psi_2 = 0,
  \end{align*}
  where $c_0 = -(-\lambda_1)^{\mu_1}$ and $c_1 = (-\lambda_1)^{\mu_1-1}$, and therefore we have a left pseudoinverse
  \[
  \psi_2
  = c_0^{-1} \left(\phi^{\mu_2} - \dots + c_1\phi \right) \circ\psi_2
  % = \phi\circ c_0^{-1} \left(\phi^{\mu_2-1} - \dots + c_1 \right) \circ\psi_2.
  \]
  This means we have a short exact sequence $0 \to \im\psi_2 \hookrightarrow M \to \coker\psi_2 \to 0$ which is split by the map $c_0^{-1} \left(\phi^{\mu_2} - \dots + c_1\phi \right)$, % which is to say $M \to \im\psi_2$ is a split surjection.

  \begin{align*}
  \chi_\phi(\phi) = (\phi - \lambda_2\id_M)^{\mu_2} (\phi - \lambda_1\id_M)^{\mu_1}
  &= \psi_2\circ\left(\phi^{\mu_1} - \dots + c_1\phi - c_0\id_M\right) \\
  &= \psi_2 c_0 - \psi_2\circ \left(\phi^{\mu_1} - \dots + c_1\phi\right) = 0,
  \end{align*}

  % $\lambda_1^{-1}A_1$ is an idempotent and by \Cref{lem:idemp} (and scaling by $\lambda_1$ for simplicity) we get $M = \ker\psi_1 \oplus \im\psi_1$.
% \devlin{But $\lambda_1\inv A_1$ is not guaranteed to be an idempotent: e.g., if $\lambda_1 = 1$ with multiplicity 2, then $A_1$ has the block $\begin{pmatrix}1 & 2 \\ 0 & 1\end{pmatrix}$ (i.e., a $2\times 2$ Jordan block squared), and it's not an idempotent.}

  Otherwise, observe that $\wpsi_j\coloneqq \psi_1\circ\dots\circ\psi_{j-1}\circ\psi_{j+1}\circ\dots\circ\psi_r $ has only one nonzero eigenvalue over $k$, namely $\widehat\lambda_j \coloneqq \prod_{i\neq j}(\lambda_j-\lambda_i)$: Using \Cref{prop:eigenvalues}, eigenvalues of $\phi$ are the same as the eigenvalues of $A$. Say $A v = \lambda_\ell v$ for $v\in M/\m M$ and apply the $k$-linear map induced by $\psi_1\circ\dots\circ\psi_r$ to get
  \begin{align}\label{eq:eigenvalues-psi}
    \prod(A-\lambda_i)v
    &= (A - \lambda_1)\cdots(A - \lambda_{r-1})(\lambda_\ell - \lambda_r)v \nonumber \\
    &= (A - \lambda_1)\cdots(\lambda_\ell - \lambda_{r-1})(\lambda_\ell - \lambda_r)v
    = \dots = \prod(\lambda_\ell - \lambda_i)v,
  \end{align}
  where the product is over $i=1,\dots,r$. In particular, since the $\lambda_i$ are distict the $k$-linear map $\widehat A_j\coloneqq (A - \lambda_1)\dots(A-\lambda_{j-1})(A - \lambda_{j+1})\dots(A - \lambda_r)$ induced by $\wpsi_j$ removes the $(\lambda_\ell-\lambda_j)$ factor from \eqref{eq:eigenvalues-psi}, hence the eigenvalue of $\widehat A_j$ corresponding to $v$ is $\prod_{i\neq j}(\lambda_j-\lambda_i)$ if $\lambda_\ell=\lambda_j$ and 0 otherwise, with multiplicities $\mu_j$ and $\mu(M) - \mu_j$ respectively.

  In particular, $\sfrac{\widehat A_j}{\widehat\lambda_j}$ is an idempotent, so by \Cref{lem:idemp} (and scaling by $\widehat\lambda_j$ for simplicity) we have $M = \ker\wpsi_j \oplus \im\wpsi_j$ and further each $\psi_j$ is a split surjection $M\to\im\psi_j = \ker\wpsi_j$. Finally, note that elements of $\ker\psi_i = \im\wpsi_i$ must be generated by columns of $\phi$ corresponding to the eigenvalue $\lambda_i$, hence $\ker\psi_i \cap \ker\psi_j = 0$ for $i\neq j$, thus by \Cref{lem:decomposition} we have the desired decomposition.
\end{proof}

%% \begin{proof}
%%   First we show that each $\psi_j$ is a split surjection $M\to\im\psi_j$, hence $M = \ker\psi_j \oplus \im\psi_j$. To see this, note that $R$-linear row and column operations on endomorphisms of $M$ correspond to pre- and post-composition by $R$-linear maps on $M$. In particular, performing row and column operations only involving elements of $R_0 = k$ on $\phi$ we may transform each $B_i$ into Jordan canonical form
%%   \begin{equation}
%%     J_i = \begin{pmatrix}
%%       J_{i,1} & 0       & \dots  & 0      \\
%%       0       & J_{i,2} & \dots  & 0      \\
%%       \vdots  & \vdots  & \ddots & \vdots \\
%%       0       & 0       & \dots  & J_{i,t},
%%     \end{pmatrix}
%%   \end{equation}
%%   where each Jordan block has entries over $k$ and a fixed eigenvalue. Suppose $J_{i,j}$ is the block with eigenvalue $\lambda_j$, then $N_i = J_{i,j} - \lambda_j$ is a nilpotent matrix whose size is the multiplicity of $\lambda-\lambda_j$ in $\chi_{B_i}(\lambda)$. Since $\mu_j$ is the sum of the multiplicities of $\lambda_j$ in each $B_i$, the matrix $N_i^{\mu_j}$ vanishes, hence $(J_i-\lambda_j)^{\mu_j}$ is a matrix with some 1s along the diagonal and 0s elsewhere. This implies that there are auto morphisms $\alpha$ and $\beta$ of $M$ (corresponding to column operations clearing the entries above and to the right of each nonzero diagonal) such that $\alpha \circ (\phi - \lambda_j\id_M)^{\mu_j} \circ \beta: M\to M$ is represented by a diagonal matrix $\psi'_j$ with 1s and 0s along the diagonal, which is a split surjection.

%%   Finally, if $\lambda_j$ and $\lambda_k$ are different then the nilpotent blocks in $J_i-\lambda_j$ and $J_i-\lambda_k$ are distinct, therefore $\ker\psi_j \cap \ker\psi_k = 0$, 
%% \end{proof}

@@

We will see via \Cref{lem:distinct} that this will result in an indecomposable decomposition of $M$ over any field.

>>>>

We will refer to the collection of eigenvalues of $B_i$, extending the field if necessary, as the ``eigenvalues of $\phi$''. Further, let $\mu_j \coloneqq \mu(\lambda_j)$ be the sum of the multiplicities of $\lambda_j$ in each $B_i$. Note that the multiplicity of $\lambda_j$ in $\chi_\phi(\lambda)$ is the sum of the multiplicities of $\lambda_j$ in each $\chi_{B_i}(\lambda)$.

@@

\begin{prop}\label{prop:split-surj}
  Let $\lambda_1,\dots,\lambda_r$ be the eigenvalues of $\phi$ over $k$ and set $\psi_j = (\phi - \lambda_j\id_M)^{\mu_j}$ and $\psi = \psi_1\circ\cdots\circ\psi_r$. Then $M$ has a direct sum decomposition
  \( M = \ker\psi_1 \oplus \cdots \oplus \ker\psi_r \oplus \im\psi. \)
\end{prop}
\begin{proof}
  First note that if $\psi_i$ is injective on $\im\psi_i$, then $\im\psi_i$ is a summand: the composition $\im\psi_i\to M\to \im\psi_i$ is injective, hence it is an isomorphism, which means the inclusion $\im\psi_i\to M$ is split. Therefore using the short exact sequence $0\to\ker\psi_i\to M\to\im\psi_i\to 0$ we get the decomposition $M = \ker\psi_i\oplus\im\psi_i$.

  To see that $\psi_i$ is injective on $\im\psi_i$, first for simplicity replace $\phi$ by $\phi-\lambda_i\id_M$ so $\lambda_i = 0$ is an eigenvalue of $\phi$ and $\psi_i = \phi^{\mu_i}$. By the Cayley--Hamilton theorem $\psi_i$ divides $\chi_\phi(\phi)$, hence
  \begin{align*}
    \chi_\phi(\phi) = \left(\phi^{\mu(M) - \mu_i} - \dots + c_1\phi - c_0\right) \circ \psi_i
    % &= \left(\phi^{\mu(M) - \mu_i} - \dots + c_1\phi\right) \circ \psi_i - c_0 \psi_i \\
    &= \left(\phi^{\mu(M) - \mu_i - 1} - \dots + c_1\right) \circ \phi \circ \psi_i - c_0 \psi_i,
  \end{align*}
  for some $c_0,c_1$ in $k$, and therefore
  \( \psi_i = c_0^{-1} \left(\phi^{\mu(M) - \mu_i - 1} - \dots + c_1\right) \circ \phi \circ \psi_i. \)
  This means that $\phi$ is injective on $\im\psi_i$, therefore $\psi_i=\phi^{\mu_i}$ is injective on $\im\psi_i$.

  Moreover, using the standard proof for vector spaces, we know $\ker\psi_i\cap\ker\psi_j = 0$ if $i\neq j$: suppose $(\phi - \lambda_i)^m(v) = 0$ for some $v\in M$ and $m$ is minimal so that $w = (\phi-\lambda_i)^{m-1}(v) \neq 0$ satisfies $(\phi-\lambda_i)(w) = 0$, but then since $(\phi - \lambda_j)(\phi - \lambda_i) = (\phi - \lambda_i)(\phi - \lambda_j)$ we get
  \[
  (\phi - \lambda_j)(w)
  = (\phi - \lambda_j)(\phi - \lambda_i)^{m-1}(v)
  =  (\phi - \lambda_i)^{m-1}(\phi - \lambda_j)(v) = 0,
  \]
  which implies
  \(
  0 = (\phi - \lambda_j)(w) = (\phi - \lambda_i + \lambda_i - \lambda_j)(w) = (\lambda_i - \lambda_j)(w),
  \)
  which is a contradiction.

  Finally, \Cref{lem:decomposition} implies the desired decomposition.
\end{proof}

@@

We will see via \Cref{lem:distinct} that this will result in an indecomposable decomposition of $M$.
